---
title: "Exploratory Analysis of NOM Selection"
author: "AnkeLi"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
data_path = ""
#"//biostat-fs2-s/users/jic034/Downloads/sonoma_cleaned.csv"
#"C:/Bre/SONOMA/Sonoma_cleaned.csv"

clean_s = read.csv(data_path, check.names = FALSE, row.names=1) 

# colnames(clean_s)
```


# Proposed covariates in multivariate logistic/multinomial

According to codebook:

-   age, continuous | `Age_Categories` (type conflict)
-   sex, binary | `Sex`
-   race, categorical | `Race`
-   hispanic ethnicity, binary | `Ethnicity`
-   english as second language, binary | mostly missing
-   charlson score, continuous | `ccioutcomes`
-   appendiceal diameter, continuous | `appy_sizeoutcomes`
-   appendicolith, binary | `appendicolithoutcomes` (1/0/2 = Yes/No/Notmentioned)
-   WBC or NLR, continuous | `wbcoutcomes`, `nlroutcomes`

```{r}

#table showing counts of eligible patients that received appendectomy (1) vs NOM (2)
table(clean_s$initial_planoutcomes)
#about 23% of eligible participants chose NOM

#table of NOM vs Appy by RaceEthnicity
table(clean_s$initial_planoutcomes, clean_s$RaceEthnicity)

#table of NOM vs Appy by Sex
table(clean_s$initial_planoutcomes, clean_s$Sex) #pretty even

#table of NOM vs Appy by Age
table(clean_s$initial_planoutcomes, clean_s$Age_Categories)



vars <- c("Age_Categories", "Sex", "Ethnicity", "appendicolithoutcomes", "ccioutcomes", "RaceEthnicity")

for (var in vars){
x <- clean_s[[var]]

tab <- table(x, useNA = "ifany")
pct <- 100 * tab / sum(tab)
lab <- paste0("Counts/frequency: ",
              paste(sprintf("%s (%d, %.1f%%)", names(tab), as.integer(tab), as.numeric(pct)),
                    collapse = ", "))

barplot(tab, main = var, sub = lab, las = 2)
}

#table of counts of Race/Ethnicity
table(clean_s$RaceEthnicity)
#RESULTS: less than 30 in AI/AN, Asian Indian, and NHPI. Will need to collapse these columns -> probably just add to other/not specified

#table of counts of age ranges
table(clean_s$Age_Categories)
# RESULTS: all categories have >30 individuals, though unbalanced as 65-74 and 75+ are way lower than other categories. Should we put them together? Maybe better to just keep continuous in the model, but show categories for descriptive data

#table of counts of ESL
table(clean_s$ESL)
table(clean_s$primary_language)
#RESULT: would be more balanced if we used ESL indicator, but worth noting in results that majority of "Other" were spanish speakers. 



```
```{r}
# ccioutcomes is better interpret in histogram i think?
vars <- c("ccioutcomes","appy_sizeoutcomes","wbcoutcomes","nlroutcomes")  # continuous

for (v in vars) {
  x <- clean_s[[v]]
  x = as.numeric(x)
  x <- x[is.finite(x)]
  op <- par(oma=c(2,0,0,0))  
  boxplot(x, main=v, ylab=v, horizontal=TRUE)
  mtext(sprintf("mean=%.2f  median=%.2f  Q1=%.2f  Q3=%.2f  min=%.2f  max=%.2f",
                mean(x), median(x),
                unname(quantile(x,.25)), unname(quantile(x,.75)),
                min(x), max(x)),
        side=1, line=.2, cex=.9, outer=TRUE)
   par(op)
}



```

# NOM with covariates

`nom_appyoutcomes`
```{r}
# barplot(clean_s$nom_appyoutcomes)



x <- clean_s[["nom_appyoutcomes"]]

tab <- table(x, useNA = "ifany")
pct <- 100 * tab / sum(tab)
lab <- paste0("Counts/frequency: ",
              paste(sprintf("%s (%d, %.1f%%)", names(tab), as.integer(tab), as.numeric(pct)),
                    collapse = ", "))

barplot(tab, main = var, sub = lab, las = 2)



```


```{r}
library(ggplot2)

# Checking if treatment selection (NOM vs Surgery) differs by Race or Age
# Using Chi-square tests to see if these associations are just random noise
race_test <- chisq.test(table(clean_s$RaceEthnicity, clean_s$initial_planoutcomes))
print(race_test)

age_test <- chisq.test(table(clean_s$Age_Categories, clean_s$initial_planoutcomes))
print(age_test)

res <- chisq.test(table(clean_s$RaceEthnicity, clean_s$initial_planoutcomes))
round(res$residuals, 2) 
# Note: Residuals > 2 or < -2 are the "red flags" for significant deviation
```

```{r}
# Showing the percentage of NOM vs Surgery for each age group
# This "fill" position makes it easy to compare the ratio rather than just counts
ggplot(clean_s, aes(x = Age_Categories, fill = as.factor(initial_planoutcomes))) +
  geom_bar(position = "fill") + 
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Treatment Selection Ratio by Age Group", 
       y = "Percentage (%)", 
       fill = "Plan (1:Appy, 2:NOM)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Visualizing the proportion of treatment by Race
ggplot(clean_s, aes(x = RaceEthnicity, fill = as.factor(initial_planoutcomes))) +
  geom_bar(position = "fill") +
  coord_flip() + # Flipping it makes those long race names easier to read
  scale_y_continuous(labels = scales::percent) +
  labs(title = "NOM vs Surgery Ratio by Race",
       x = "Race/Ethnicity", y = "Percentage (%)", 
       fill = "Plan (1:Appy, 2:NOM)") +
  theme_minimal()

# plotting those residuals we calculated earlier
res_df <- as.data.frame(res$residuals)
colnames(res_df) <- c("Race", "Treatment", "Residual")

ggplot(res_df, aes(x = Race, y = Residual, fill = Treatment)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", color = "red") + # The +/- 2 threshold
  coord_flip() +
  labs(title = "Where the Selection Bias Is: Residuals by Race",
       subtitle = "Bars crossing the red lines are statistically significant",
       y = "Standardized Residuals", x = "") +
  scale_fill_discrete(labels = c("1: Appendectomy", "2: NOM")) +
  theme_minimal()

```

```{r}
# Prepping the labels to be "Yes/No" or "NOM/Appy" 
clean_s$initial_plan_factor <- factor(clean_s$initial_planoutcomes, 
                                     levels = c(1, 2), 
                                     labels = c("Appendectomy", "NOM"))

#Plot the proportion of two treatments by Charlson Comorbidity Index (CCI)
p_cci <- ggplot(clean_s, aes((x = ccioutcomes), fill = initial_plan_factor)) +
  geom_bar(position = "fill") + 
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Treatment Selection by Comorbidity (Cleaned CCI)", 
       x = "Charlson Comorbidity Index (Rounded to nearest integer)", 
       y = "Proportion (%)", 
       fill = "Treatment") +
  theme_minimal()

print(p_cci)


# Fecalith (Appendicolith) analysis
clean_s$lith_factor <- factor(clean_s$appendicolithoutcomes, 
                              levels = c(1, 0, 2), 
                              labels = c("Yes", "No", "Not Mentioned"))

p_lith <- ggplot(clean_s, aes(x = lith_factor, fill = initial_plan_factor)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Fecalith Impact on Selection", 
       x = "Appendicolith Present?", y = "Proportion", fill = "Treatment") +
  theme_minimal()


print(p_lith)

```

```{r}
library(dplyr)
library(ggplot2)
library(gridExtra)


analysis_data <- clean_s %>%
  filter(!is.na(initial_plan_factor)) %>% 
  mutate(
    wbc_val = as.numeric(as.character(wbcoutcomes)),
    nlr_val = as.numeric(as.character(nlroutcomes)),
    dia_val = as.numeric(as.character(appy_sizeoutcomes))
  )

# Simple Pairwise Table 
# This calculates stats for each variable independently based on its available N
clinical_summary <- analysis_data %>%
  group_by(initial_plan_factor) %>%
  summarise(
    # WBC: Total N = 1281 (Appy) / 376 (NOM)
    n_WBC      = sum(!is.na(wbc_val)),
    Median_WBC = median(wbc_val, na.rm = TRUE),
    
    # NLR  has higher missingness
    n_NLR      = sum(!is.na(nlr_val)),
    Median_NLR = median(nlr_val, na.rm = TRUE),
    
    # Diameter  
    n_Dia      = sum(!is.na(dia_val)),
    Median_Dia = median(dia_val, na.rm = TRUE)
  )

print("Final Summary Table (Filtered for missing outcomes):")
print(clinical_summary)

#  Visualizations 
plot_clinical_distribution <- function(data, variable, plot_title, y_label, use_log_scale = FALSE) {
  p <- ggplot(data, aes(x = initial_plan_factor, y = .data[[variable]], fill = initial_plan_factor)) +
    geom_violin(alpha = 0.3, color = NA, trim = TRUE, na.rm = TRUE) + 
    geom_boxplot(width = 0.2, outlier.size = 1, alpha = 0.6, na.rm = TRUE) +
    labs(title = plot_title, x = "", y = y_label) +
    theme_minimal() +
    theme(legend.position = "none", plot.title = element_text(face = "bold"))
  
  if(use_log_scale) {
    p <- p + scale_y_log10()
  }
  return(p)
}

p_wbc <- plot_clinical_distribution(analysis_data, "wbc_val", "WBC Distribution", "10^9/L (Log)", use_log_scale = TRUE)
p_nlr <- plot_clinical_distribution(analysis_data, "nlr_val", "NLR Distribution", "Ratio (Log)", use_log_scale = TRUE)
p_dia <- plot_clinical_distribution(analysis_data, "dia_val", "Appendiceal Diameter", "mm")

grid.arrange(p_wbc, p_nlr, p_dia, ncol = 3)
```



```{r}
#Wilcoxon test see NOM and surgery group difference
wbc_p_value <- wilcox.test(wbc_val ~ initial_plan_factor, data = analysis_data)$p.value
nlr_p_value <- wilcox.test(nlr_val ~ initial_plan_factor, data = analysis_data)$p.value
dia_p_value <- wilcox.test(dia_val ~ initial_plan_factor, data = analysis_data)$p.value

test_results <- data.frame(
  Variable = c("WBC Count", "NLR Ratio", "Appendiceal Diameter"),
  P_Value = c(wbc_p_value, nlr_p_value, dia_p_value),
  Significance = ifelse(c(wbc_p_value, nlr_p_value, dia_p_value) < 0.05, "Significant", "Not Significant")
)

print("Simple Statistical Test Results (Appendectomy vs NOM):")
print(test_results)
```