---
title: "eda"
output: html_document
date: "2026-01-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#data_path = "//biostat-fs2-s/users/jic034/Downloads/sonoma_cleaned.csv"
data_path = "C:/Bre/SONOMA/Sonoma_cleaned.csv"

clean_s = read.csv(data_path, check.names = FALSE, row.names=1) 

# colnames(clean_s)
```

# Covariates in multivariate logistic/multinomial

According to codebook:

-   age, continuous | `Age_Categories` (type conflict)
-   sex, binary | `Sex`
-   race, categorical | `Race`
-   hispanic ethnicity, binary | `Ethnicity`
-   english as second language, binary | mostly missing
-   charlson score, continuous | `ccioutcomes`
-   appendiceal diameter, continuous | `appy_sizeoutcomes`
-   appendicolith, binary | `appendicolithoutcomes` (1/0/2 = Yes/No/Notmentioned)
-   WBC or NLR, continuous | `wbcoutcomes`, `nlroutcomes`

```{r}

#table showing counts of eligible patients that received appendectomy (1) vs NOM (2)
table(clean_s$initial_planoutcomes)
#about 23% of eligible participants chose NOM

#table of NOM vs Appy by RaceEthnicity
table(clean_s$initial_planoutcomes, clean_s$RaceEthnicity)

#table of NOM vs Appy by Sex
table(clean_s$initial_planoutcomes, clean_s$Sex) #pretty even

#table of NOM vs Appy by Age
table(clean_s$initial_planoutcomes, clean_s$Age_Categories)



vars <- c("Age_Categories", "Sex", "Ethnicity", "appendicolithoutcomes", "ccioutcomes", "RaceEthnicity")

for (var in vars){
x <- clean_s[[var]]

tab <- table(x, useNA = "ifany")
pct <- 100 * tab / sum(tab)
lab <- paste0("Counts/frequency: ",
              paste(sprintf("%s (%d, %.1f%%)", names(tab), as.integer(tab), as.numeric(pct)),
                    collapse = ", "))

barplot(tab, main = var, sub = lab, las = 2)
}

#table of counts of Race/Ethnicity
table(clean_s$RaceEthnicity)
#RESULTS: less than 30 in AI/AN, Asian Indian, and NHPI. Will need to collapse these columns -> probably just add to other/not specified

#table of counts of age ranges
table(clean_s$Age_Categories)
# RESULTS: all categories have >30 individuals, though unbalanced as 65-74 and 75+ are way lower than other categories. Should we put them together? Maybe better to just keep continuous in the model, but show categories for descriptive data

#table of counts of ESL
table(clean_s$ESL)
table(clean_s$primary_language)
#RESULT: would be more balanced if we used ESL indicator, but worth noting in results that majority of "Other" were spanish speakers. 



```
```{r}
# ccioutcomes is better interpret in histogram i think?
vars <- c("ccioutcomes","appy_sizeoutcomes","wbcoutcomes","nlroutcomes")  # continuous

for (v in vars) {
  x <- clean_s[[v]]
  x = as.numeric(x)
  x <- x[is.finite(x)]
  op <- par(oma=c(2,0,0,0))  
  boxplot(x, main=v, ylab=v, horizontal=TRUE)
  mtext(sprintf("mean=%.2f  median=%.2f  Q1=%.2f  Q3=%.2f  min=%.2f  max=%.2f",
                mean(x), median(x),
                unname(quantile(x,.25)), unname(quantile(x,.75)),
                min(x), max(x)),
        side=1, line=.2, cex=.9, outer=TRUE)
   par(op)
}



```



# NOM with covariates

`nom_appyoutcomes`
```{r}
# barplot(clean_s$nom_appyoutcomes)



x <- clean_s[["nom_appyoutcomes"]]

tab <- table(x, useNA = "ifany")
pct <- 100 * tab / sum(tab)
lab <- paste0("Counts/frequency: ",
              paste(sprintf("%s (%d, %.1f%%)", names(tab), as.integer(tab), as.numeric(pct)),
                    collapse = ", "))

barplot(tab, main = var, sub = lab, las = 2)



```
# EDA for NOM Patients
```{r}
library(gmodels)
NOM_patients <- clean_s[clean_s$initial_planoutcomes == 2,] 

NOM_patients <- NOM_patients[!is.na(NOM_patients$time_of_dxoutcomes),]

#setting NA values to 0 for censored, creating status variable for survival
NOM_patients$status <- ifelse(
  is.na(NOM_patients$interval_typeoutcomes),
  0,   # censored
  NOM_patients$interval_typeoutcomes
)

table(NOM_patients$status)
#Post NOM, 41 received elective appendectomy, 89 received Urgent/Emergent appendectomy, and 251 received neither. Do we want to adjust this so its only people who received appy after 1 year?

#Crosstable of RaceEthnicity by Status
CrossTable(NOM_patients$RaceEthnicity, NOM_patients$status)
#some groups like Asian, Multiracial, etc. only have 1 or 0 patients in certain outcome groups

#Crosstabulation of abnormal vital indicator by status
CrossTable(NOM_patients$abnormal_vital_indicator, NOM_patients$status)

#Crosstabulation of appendicolith and status
CrossTable(NOM_patients$appendicolithoutcomes, NOM_patients$status)



library(nnet) #testing to see if there is any association b/w status and appendicolith
#multi_mo <- multinom(status ~ appendicolithoutcomes, data = NOM_patients ,model=TRUE)
#summary(multi_mo)
#z <- summary(multi_mo)$coefficients/summary(multi_mo)$standard.errors
#z
#p <- (1 - pnorm(abs(z), 0, 1)) * 2
#p


#check for multicollinearity b/w variables?
#check for linear relationship b/w continuous variables and log-odds of outcome?
#potentially check for outliers of continuous variables?

```
