```{r}
library(ggplot2)
library(dplyr)
library(lubridate)
library(knitr)
library(tidyr)
library(gtsummary)
library(kableExtra)
library(webshot2)

```


# Load Start Dates
```{r}
start_dates = read.csv("SONOMA-Data/TRIADPhase2_Site_Start_Date_V5_20231215.csv")
start_dates = start_dates %>% select(-site_num)
start_dates$start_date = as.POSIXct(start_dates$start_date)

nrow(data[data$choice == "NOM",])
```


# Load and clean data
```{r}
data = read.csv("SONOMA-Data/sonoma_cleaned_2_20260202.csv")

data$firstdxoutcomes = as.POSIXct(data$firstdxoutcomes)
data$initial_planoutcomes = data$initial_planoutcomes - 1
data$choice = factor(data$initial_planoutcomes, levels = c(0,1), labels = c("Surgery", "NOM" ))
data = merge(x = data, y = start_dates, by = "site_name", all.x = TRUE)

data <- data %>%
  filter(!is.na(choice) ) %>%
  filter(site_name != "Lyndon B Johnson Hospital") %>%
  relocate(RecordID, 
           site_name, 
           initial_planoutcomes, 
           choice,
           start_date,
           firstdxoutcomes) %>%
  filter(firstdxoutcomes > as.Date("2022-12-01") & firstdxoutcomes < as.Date("2024-08-30"))

```


# Initial Plot, raw data
```{r, fig.width=8, fig.height=6}
raw_plot = ggplot(data = data, aes(x = as.Date(firstdxoutcomes), y = choice)) + 
  geom_jitter(alpha = .2, size = 1, width = .1, height = .01) + 
  geom_vline(aes(xintercept = as.Date(start_date), color = "AppyOrNot"), size = .75) +
  labs(color = "") +
  xlab("Date of Index") +
  ylab("Choice at Index") + 
  geom_text(
    data = start_dates[start_dates$site_name %in% unique(data$site_name), ],
    aes(
      label = format(start_date, "%B %d %Y"),
      x = as.Date(start_date) %m+% days(140), 
      y = 1.2
    ),
    vjust = -0.6,
    size = 3.2,
    color = "red"
  ) +
  facet_wrap(~site_name) +
  theme_bw() +
  labs(title = "NOM vs. Surgery Selections Over Time") + 
scale_x_date(
    breaks = 
      as.Date(c(
        "2023-4-01",
        "2024-4-01"
      ))
      ,
    labels = function(x) format(x, "%B %d %Y")  # e.g., "January 2020"
  ) + 
  theme(strip.text = element_text(size = 7))




ggsave("/home/users/sherold/Documents/SONOMA-figures/raw_plot.png", plot = raw_plot, width = 12, height = 6)

```


# Introduce Pre/Post periods and summarize on UNBINNED DATA
```{r}
ts_unbinned = data %>%
  group_by(site_name, RecordID, firstdxoutcomes) %>%
  summarise(
    n_TOTAL = n(),
    n_NOM = sum(initial_planoutcomes == 1, na.rm = TRUE),
    n_SURGERY = sum(initial_planoutcomes == 0, na.rm = TRUE),
    NOM_rate = n_NOM/n_TOTAL,
    start_date = as.Date(start_date),
    .groups = "drop"
  )
  #filter(firstdxoutcomes > as.Date("2022-12-01") & firstdxoutcomes < as.Date("2024-08-30"))
  
ts_unbinned = ts_unbinned %>% mutate(post = case_when(firstdxoutcomes >= start_date ~ 1,
                               firstdxoutcomes < start_date ~ 0))  
```


```{r}
# Get Site totals and rates
site_summary <- ts_unbinned %>%
  group_by(site_name, post) %>%
  summarise(
    N = sum(n_TOTAL),
    NOM_Count = sum(n_NOM),
    NOM_rate = sum(n_NOM) / sum(n_TOTAL),
    .groups = "drop"
  ) %>%
  mutate(
    Period = if_else(post == 1, "Post", "Pre")
  ) %>%
  select(site_name, Period, N, NOM_Count, NOM_rate)

site_summary_wide <- site_summary %>%
  pivot_wider(
    names_from = Period,
    values_from = c(N, NOM_Count, NOM_rate)
  )

# Add overall (All sites) row
site_summary_wide <- bind_rows(
  site_summary_wide,
  site_summary_wide %>%
    summarise(
      site_name = "All",
      N_Pre = sum(N_Pre, na.rm = TRUE),
      N_Post = sum(N_Post, na.rm = TRUE),
      NOM_Count_Pre = sum(NOM_Count_Pre, na.rm = TRUE),
      NOM_Count_Post = sum(NOM_Count_Post, na.rm = TRUE),
      NOM_rate_Pre = NOM_Count_Pre / N_Pre,
      NOM_rate_Post = NOM_Count_Post / N_Post
    )
)

# Add rate difference + format for display
site_summary_wide <- site_summary_wide %>%
  mutate(
    NOM_Rate_Diff = NOM_rate_Post - NOM_rate_Pre,
    across(starts_with("NOM_rate"), ~ round(. * 100, 1)),
  ) %>%
  rename(
    "Site" = site_name,
    "Events Pre" = N_Pre,
    "Events Post" = N_Post,
    "NOM Count Pre"= NOM_Count_Pre,
    "NOM Count Post" = NOM_Count_Post,
    "NOM Rate Pre" = NOM_rate_Pre,
    "NOM Rate Post" = NOM_rate_Post,
    "Change in NOM Rate" = NOM_Rate_Diff
  )
                                   
```

# Plot TS Table 1
```{r}
site_summary_wide %>%
  mutate(across((ncol(.)-2):ncol(.), ~ paste0(.x, "%"))) %>%
  kbl(
    align = rep("l", ncol(.))
  ) %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>% row_spec(nrow(site_summary_wide)-1,
         extra_css = "border-bottom: 2px solid black;")

```

# Monthly Rates
```{r}
ts_month = data %>%
  mutate(month = as.Date(floor_date(firstdxoutcomes, unit = "month"))) %>%
  group_by(site_name, month) %>%
  summarise(
    n_TOTAL = n(),
    n_NOM = sum(initial_planoutcomes == 1, na.rm = TRUE),
    n_SURGERY = sum(initial_planoutcomes == 0, na.rm = TRUE),
    NOM_rate = n_NOM/n_TOTAL,
    start_date = first(as.Date(start_date)),
  ) %>%
  filter(month > as.Date("2022-12-01") & month < as.Date("2024-08-30"))%>% 
                               mutate(month = month + 14)
```


```{r, fig.width=8, fig.height=6}
rate_plot = ggplot(data = ts_month, 
       aes(x = month, y = NOM_rate, group = site_name)) + 
  geom_point(size = 1,) + 
  geom_line(size = .75, alpha = .6) +
  geom_vline(aes(xintercept = as.Date(start_date), color = "AppyOrNot"), size = .75) +
  labs(color = "") +
  xlab("Date") +
  ylab("Montly NOM Rate") + 

  facet_wrap(~site_name) +
  theme_bw() +
  labs(title = "Monthly NOM Rate Over Time") + 
scale_x_date(
    breaks = 
      as.Date(c(
        "2023-4-01",
        "2024-4-01"
      ))
      ,
    labels = function(x) format(x, "%B %d %Y")  # e.g., "January 2020"
  ) + 
  theme(strip.text = element_text(size = 7))




ggsave("/home/users/sherold/Documents/SONOMA-figures/rate_plot.png", plot = rate_plot, width = 12, height = 6)

```
```



# Introduce Click Rate

```{r}
clicks_raw = read.csv("SONOMA-Data/Site Tally(Clicks per month)_cleaned.csv")
month_cols <- c("Oct.23","Nov.23","Dec.23","Jan.24","Feb.24","Mar.24",
                "Apr.24","May.24","Jun.24","Jul.24","Aug.24","Sep.24", 
                "Oct.24", "Nov.24", "Dec.24", "Jan.25", "Feb.25", "Mar.25")

clicks = clicks_raw %>%
  select(-c(Date.TTT.Completed.Site.Launch, Total, average.for.2024)) %>% 
  mutate(across(all_of(month_cols), ~ as.numeric(.))) %>% 
  pivot_longer(
    cols = all_of(month_cols),
    names_to = "month",
    values_to = "clicks"
  ) %>%
  mutate(
    month = as.Date(paste0("01-", month), format = "%d-%b.%y")
  ) 

ts_clicks = merge(x = ts_month, y = clicks, by = c("site_name", "month"), all.x = TRUE) %>% 
    mutate(across(everything(), ~ replace_na(., 0)))

ts_clicks = ts_clicks %>% mutate(post = case_when(month >= start_date ~ 1,
                               month < start_date ~ 0))  %>% 
                               mutate(month = month + 14)

```
```

```{r, fig.width=8, fig.height=6}
click_plot = ggplot(data = ts_clicks, 
       aes(x = month, y = clicks, group = site_name)) + 
  geom_point(size = 1,) + 
  geom_line(size = .75, alpha = .6) +
  geom_vline(aes(xintercept = as.Date(start_date), color = "AppyOrNot"), size = .75) +
  labs(color = "") +
  xlab("Date") +
  ylab("Monthly AppyOrNot Clicks") + 

  facet_wrap(~site_name) +
  theme_bw() +
  labs(title = "Monthly AppyOrNot Clicks") + 
scale_x_date(
    breaks = 
      as.Date(c(
        "2023-4-01",
        "2024-4-01"
      ))
      ,
    labels = function(x) format(x, "%B %d %Y")  # e.g., "January 2020"
  ) + 
  theme(strip.text = element_text(size = 7))




ggsave("/home/users/sherold/Documents/SONOMA-figures/click_plot.png", plot = click_plot, width = 12, height = 6)
```

```{r}
click_nom_compare = ggplot(data = ts_clicks, aes(x = clicks, y = as.factor(n_NOM))) + 
  geom_point(alpha = .25) +
  xlab("Monthly AppyOrNot Clicks") +
  ylab("Monthly NOM Count") +
  facet_wrap(~site_name) +
  theme_bw() + 
  theme(strip.text = element_text(size = 5)) +
  labs(title = "Monthly AppyOrNot Clicks and NOM Count")

ggsave("/home/users/sherold/Documents/SONOMA-figures/click_nom_compare.png", plot = click_nom_compare, width = 6, height = 6)
```


