---
title: "NOM Outcomes - EDA"
output: html_document
date: "2026-01-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(ggplot2)
```

```{r, echo = FALSE, message = FALSE}
#data_path = "//biostat-fs2-s/users/jic034/Downloads/sonoma_cleaned.csv"
data_path = "C:/Bre/SONOMA/Sonoma_cleaned.csv"

clean_s = read.csv(data_path, check.names = FALSE, row.names=1) 

# colnames(clean_s)
```


The study dataset was subsetted to only individuals who received NOM at index. Of the 381 who received NOM, 41 (11%) received elective appendectomy and 89 (24%) received an urgent/emergent appendectomy. Limiting this to just one year follow-up from index, 39 (10%) received elective appendectomy and 80 (21%) received an urgent/emergent appendectomy.
```{r, echo = FALSE, message = FALSE}

library(gmodels)
NOM_patients <- clean_s[clean_s$initial_planoutcomes == 2,] 

NOM_patients <- NOM_patients[!is.na(NOM_patients$time_of_dxoutcomes),]

#setting NA values to 0 for censored, creating status variable for survival
NOM_patients$status <- ifelse(
  is.na(NOM_patients$interval_typeoutcomes),
  0,   # censored
  NOM_patients$interval_typeoutcomes
)

 

#Subsetting this further to 1 year post index

NOM_patients <- NOM_patients[!is.na(NOM_patients$time_of_dxoutcomes),] #removing NA values for date/time of patients dx of appendicitis
NOM_patients$time_of_dxoutcomes <- gsub(" .*","", NOM_patients$time_of_dxoutcomes) #removing time part of date/time


#converting date_index to date object
NOM_patients$time_of_dxoutcomes <- as.Date(
  NOM_patients$time_of_dxoutcomes,
  format = "%m/%d/%Y"
)


#converting date of appendectomy after NOM to date object, converting blanks to NAs
NOM_patients$appy_int_dateoutcomes[
  NOM_patients$appy_int_dateoutcomes == ""
] <- NA

NOM_patients$appy_int_dateoutcomes <- as.Date(
  NOM_patients$appy_int_dateoutcomes,
  format = "%m/%d/%Y"
)

#creating variable that is time in days from NOM decision to appy
NOM_patients$time_to_appy <- NOM_patients$appy_int_dateoutcomes - NOM_patients$time_of_dxoutcomes

NOM_patients$status365 <- ifelse(
  NOM_patients$time_to_appy > 365, 
  0, #administratively censored beyond 1 year
  NOM_patients$status
)
#setting NA values to 0 for censored,
NOM_patients$status365 <- ifelse(
  is.na(NOM_patients$status365),
  0,   # censored
  NOM_patients$status365
)

NOM_patients$status <- factor(NOM_patients$status, 0:2, c("None", "Elective", "Urgent/Emergent"))

table(NOM_patients$status)
#Post NOM, 41 received elective appendectomy, 89 received Urgent/Emergent appendectomy, and 251 received neither.

NOM_patients$status365 <- factor(NOM_patients$status365, 0:2, c("None", "Elective", "Urgent/Emergent"))

table(NOM_patients$status365)


```


RaceEthnicity has the most categories out of our potential covariates. When cross-tabulating these categories against post-NOM appendectomy status, there are many sparse or empty cells. If we were to use RaceEthnicity in a multivariate model, we would have to collapse some categories to prevent quasi-complete separation. Due to the sparseness, Fisher's exact test was used to test for an association. A p-value of 0.4608 indicates there is not a statistically significant relationship between Race/Ethnicity and post-NOM appendectomy status.
```{r, echo = FALSE}

#Crosstable of RaceEthnicity by Status
CrossTable(NOM_patients$RaceEthnicity, NOM_patients$status)


cont_table <- table(NOM_patients$status, NOM_patients$RaceEthnicity)
fisher.test(cont_table, workspace = 1e7, simulate.p.value = TRUE)


#check for multicollinearity b/w variables?
#check for linear relationship b/w continuous variables and log-odds of outcome?
#potentially check for outliers of continuous variables?

```

One-way ANOVA tests were used to test the association between post-NOM appendectomy status and continuous variables like age, WBC, and appendix diameter. All of these were found to not be significantly associated with post-NOM appendectomy status. Due to some sparse cells, Fisher's exact test was used to test the association between CCI and post-NOM appendectomy status and this test was also not statistically significant. Lastly, chi-squared tests were used to test the association between post-NOM appendectomy status and categorical variables like appendicolith presence and sex. Both these showed a not statistically significant relationship. 
```{r, echo = FALSE, message = FALSE}


library(ggstatsplot)
#Using one-way ANOVA to test for statistically significant association b/w a categorical outcome and continuous dependent variable. 

#CCI is nominal 
#wbc outcomes

#Testing Age
ggplot(data = NOM_patients, aes(x = ageoutcomes, y = status365)) +
  geom_boxplot() +
  labs(y = "Post NOM Appendectomy Status", x = "Age")


anova_model <- aov(ageoutcomes ~ status, data = NOM_patients)
summary(anova_model)
#not significant

ggplot(data = NOM_patients, aes(x = wbcoutcomes, y = status365)) +
  geom_boxplot() +
  labs(y = "Post NOM Appendectomy Status", x = "WBC")
anova_model2 <- aov(wbcoutcomes ~ status, data = NOM_patients)
summary(anova_model2)
#not significant

#CrossTable of CCI and status to determine if should use Fisher's exact test or chi-squared
NOM_patients$ccioutcomes <- as.factor(NOM_patients$ccioutcomes)
CrossTable(NOM_patients$ccioutcomes, NOM_patients$status)
contingency_table <- table(NOM_patients$status, NOM_patients$ccioutcomes)
fisher.test(contingency_table, workspace = 1e7)


#fisher test shows no association b/w CCI and either elective/urgent appendectomy

#Testing association of appy size and status
ggplot(data = NOM_patients, aes(x = appy_sizeoutcomes, y = status365)) +
  geom_boxplot() +
  labs(y = "Post NOM Appendectomy Status", x = "Appendix Diameter")
anova_model3 <- aov(appy_sizeoutcomes ~ status, data = NOM_patients)
summary(anova_model3)
#92/354 missing appendix size at index


#Testing association of appendicolith and status
ggbarstats( #completes chi-squared p-val and graphs
  data = NOM_patients,
  x = appendicolithoutcomes,
  y = status, 
  results.subtitle = TRUE, 
  proportion.test = FALSE, 
  bf.message = FALSE, 
)

#Testing Sex
ggbarstats( #completes chi-squared p-val and graphs
  data = NOM_patients,
  x = Sex,
  y = status, 
  results.subtitle = TRUE, 
  proportion.test = FALSE, 
  bf.message = FALSE, 
)

```

